Efficiently summarising relationships in large samples: a general duality between statistics of genealogies and genomes

### 2019-09-11 treestats_paper.tex
a few words in the disco

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \ln n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$. Therefore, the overall running time
of Algorithms~\algref{B} is of order $(n + \rho \log n)\log_2 n$, which is
\[
    O\left(n \log n + \rho \log^2 n \right)
\]
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

\begin{figure}

% Run the script illustrate_algorithm.py to generate these pdfs and
% the latex tables below.
\begin{center}
\begin{tabular}{ccc}
\includegraphics[width=4cm]{figures/tree_0_init.pdf} & &
\includegraphics[width=4cm]{figures/tree_0_out_0.pdf}
\\

@@

\krt{need to sort out the big-O issue here}
We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
three orders of magnitude faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids.
The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
and does not include the time required to build the genotype matrix, nor the memory required to store it.
For the largest samples size of $n=10^6$,
the matrix is approximately 200 gigabytes, and thus not practical on many systems.
The largest tree sequence simulated required less than one gigabyte of memory.

@@

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$. For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \texttt{tskit}, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \krt{need cite for this}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the 
        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
        \label{fig:stats_performance}
    }
\end{figure}

@@

In this paper,
we consider only so-called \emph{single site} statistics
(which we will define concretely shortly).
Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
such as linkage disequlibrium,
are straightforward, and are planned for future work.
Haplotype-based statistics may require a different class of algorithms.

>>>>

To analyse Algorithm~\algref{B} we will assume that the tree sequence has
been simulated under the standard coalescent with a sample of size
$n$ and population scaled recombination rate $\rho = 4 N_e r$,
where $r$ is the mean number of recombinations per chromosome per unit of time.
Under these conditions, there are $O(n + \rho \ln n)$ edges in the tree sequence
\citep{kelleher2016efficient}.  Clearly, each edge is examined
exactly once in steps \algref{B6} and \algref{B7} in order to create
the trees, and at most once in steps \algref{B3} and \algref{B4} (we do
not remove the edges for the final tree). Additionally, each edge that
we insert or remove after the initial building of the tree
may incur the cost of traversing up the tree as
far as the root in steps \algref{B5} and \algref{B8}.
As coalescent genealogies are asymptotically
balanced \citep{li2013coalescent}, the expected number of nodes
on the path to root is $\log_2 n$.
Therefore, the overall running time
of Algorithms~\algref{B} is of order $n + \rho \log(n) \log_2(n)$, which is
\[
    O\left(n + \rho \log(n)^2 \right)
\]
Computing a site statistic will have the same complexity,
as the same internal state is maintained,
and the number of segregating sites is proportional to the number of edges
(with constant equal to the ratio of mutation to recombination rate).
The asymptotic analysis here assumes that we traverse upwards to root
for each edge, but this is not the case. Edges are removed in
oldest-first order, guaranteeing that the deepest node any subtree being
modified is processed first. Therefore, this subtree will be disconnected from the
main tree, ensuring that we traverse upwards all the way to root only once.
Similarly, edges are inserted in youngest-first order, ensuring that subtrees are only
reconnected to the main tree when they are fully built.

\begin{figure}
    % Run the script illustrate_algorithm.py to generate these pdfs and
    % the latex tables below.
    \begin{center}
    \begin{tabular}{ccc}
    \includegraphics[width=4cm]{figures/tree_0_init.pdf} & &
    \includegraphics[width=4cm]{figures/tree_0_out_0.pdf}
    \\

@@

\krt{need to sort out the big-O issue here}
We used coalescent simulations to compare the performance of calculating Tajima's \citeyearpar{Tajima1989-de} $D$
statistic from tree sequences to calculating it from integer matrices containing genotypes at all variable positions.
\autoref{fig:stats_performance} shows that the tree sequence methods implemented in \texttt{tskit} processes variant data
substantially faster than matrix-based approaches once the sample size is above $n \approx 500$ haploids --
three times faster for one thousand haplotypes,
growing to nearly three orders of magnitude faster for one million haplotypes.
The expected number of mutations for each replicate is $10^4 \times \sum_{i=1}^{n-1}\frac{1}{i}$, and thus ranges from
$\approx 32,000$ to $\approx 138,000$ \citep{Watterson1975-ej}.
\autoref{fig:stats_performance} only considers the time required to calculate the statistic,
and does not include the time required to build the genotype matrix, nor the memory required to store it.
For the largest samples size of $n=10^6$,
the matrix is approximately 200 gigabytes, and thus not practical on many systems.
The largest tree sequence simulated required less than one gigabyte of memory.

@@

\begin{figure}
    \centering
    \includegraphics{tskit_stat_benchmarks/benchmarks_without_copy_longer_genome.pdf}
    \caption{Comparison of time required to compute site statistics
        between matrix and tree based methods. For each sample size, a single replicate
        was obtained using \texttt{msprime} with scaled neutral mutation rate $\theta = 4N_e\mu$ and
        scaled recombination rate $\rho = 4N_er$ both equal to $10^4$ and $N_e = 10^4$. For each replicate, Tajima's $D$
        \citep{Tajima1989-de} statistic was calculated using \tskit, \texttt{libsequence}
        \citep{Thornton2003-wj}, and \texttt{scikit-allel} \citep{miles2017scikit}.
        The solid green line shows the result of fitting the timing data to a model based on the expected complexity of the 
        algorithm used by \texttt{tskit}. \krt{need to revisit the correctness of this}
        \label{fig:stats_performance}
    }
\end{figure}

@@

% this stuff is faster
The most direct effect of the methods presented here might be on efficiency of computation
with large genomic datasets --
the python library \tskit{} provides fast, rigorously tested implementations
of many standard population genetic statistics, with more on the way.
The logarithmic scaling of storage and computation time with sample size
could bring much more computation-intensive methods within reach on modern large genomic datasets
of humans, 
and helps make detailed investigation of whole populations in evolutionary simulations feasible.

% but the duality thing is maybe more long-term exciting
However, the focus of this paper was not on computation of standard population genetics statistics,
but on the relationship of these statistics to the underlying trees.
We have used essentially the same general framework
to define two additional classes of statistics -- Branch and Node statistics --
that well-defined relationships to the Site statistics computable from genome sequence.
% describing trees is cool
At least, this gives us descriptive tools to directly describe the genealogical trees
of tree sequences obtained by simulation.
% and we can do it with real data now
However, new methods to infer tree sequences allow us to apply these methods to real data.
% branch stats are less noisy!
Given the true tree sequence,
Branch statistics are less-noisy versions of the standard Site statistics,
so methods that use Branch statistics may show increased power and precision
over the corresponding genotype-only methods.
We have shown that this is true in simulated data,
but it is not yet known how much additional noise the process of tree sequence inference will add.
% and we get the time dimension
Furthermore, tree sequences add a \emph{time} dimension to genotype data,
which these descriptive methods open up many possibilities for utilizing and visualizing of.

\paragraph{Future work}
In this paper,
we have considered only so-called additive, single site statistics.
Extensions to statistics involving the pairwise joint distribution of genotypes across sites,
such as linkage disequlibrium are planned for future work.
Haplotype-based statistics may require a different class of algorithms.
